D3064R0: How to Avoid OOTA Without Really Trying

### 2024-03-29 ootavacuous.tex
Make instruction cost commentary consistent
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Final (?) edits for appendices
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Next, definitions:
\begin{description}
\item[coe]
	Coherence-order (external), which connects a store to any other store
	to the same object that comes later in the object's
	modification order, i.e., that overwrites either the first store's
	value or some later value.
\item[fre]
	From-read (external), which connects a load to any store to
	the same object that overwrites either the value that the load
	returned or some later value.
\item[rfe]
	Reads-from (external), which connects a store to any load from
	the same object that returns the value stored.
	(Note: This means that the load retrieves the information
	written by the store, as opposed to retrieving the information
	from a different store that happened to write the same value.)
\end{description}

@@

Figure~\ref{fig:Propagation Delay and Temporal Properties of coe, fre, and rfe}
shows how propagation delay explains the temporal properties of coe,
fre, and rfe.
The upper portion of the figure shows an atemporal coe, in which
CPU~0 stores 1 to \co{x} after CPU~3 stores 2 but
the modification order of \co{x} is decided after the fact.
In this case, the value of 2 from CPU~3's store overwrites the value
of 1 from CPU~0's store, despite the fact that CPU~0's store
happened later in global time.
One way this could happen is if the cache line containing
\co{x} arrived at CPU~3 before arriving at CPU~0.
A more fanciful explanation is that both values arrived at an
interconnect at the same time, and the interconnect made an
arbitrary choice between the two.
Either way, the end result is that an earlier store can overwrite
a later store.

@@

The \co{-funsigned-char} causes the \co{char} type to be unsigned, which
overrides per-architecture defaults, some of which treat \co{char}
as signed and others as \co{unsigned}.
This choice prevents a class of bugs, and also allows the kernel to make
reliable use of the uppermost bit of variables of type \co{char}.
It also affects the definition of ``semantic dependency'' by changing
the arithmetic proporties of this type.
In theory, the standard could have specified the signedness of \co{char}
but the variety of existing practice in the 1980s prevented this.
Plus machine instructions were much more expensive back then, so the
signedness of \co{char} was much more of a performance concern than
it is today.

@@

However, this would mean that programs compiled in this way would not
be guaranteed to run correctly on weakly ordered systems that lack
ECC (or that lack precise exceptions).
In addition, this change is more strict than necessary, forbidding
optimizations that tooling could in fact analyze.
Therefore, the next section looks at standardizing a less severe
restriction.

@@

This suggests the addition of a \co{memory_order_load_store} member to
the \co{memory_order} enumeration, a topic taken up by the next section.

@@

But if \co{x} is a nonvolatile atomic, there are no worries about
data races.
In theory, if the \co{do_something_with()} call was not present
the compiler could still make the above transformation without
worry, especially given that the value loaded from \co{x} is discarded in
the case where \co{limit} is less than or equal to zero.
Except that the data-race worry is replaced by a cache-miss worry.
After all, optimizations are supposed to speed things up.
A compiler might be wise to adhere to the
same restriction that avoids data races for loads from non-atomic
variables in order to avoid potentially expensive cache misses.

@@

On line~14 of this listing, the computation \co{1 / (r2 <= 0)}
will yield one if \co{r2} is not positive and will result in UB
otherwise (a divide-by-zero error).
It may not be obvious, but the program is in fact UB-free---there are
no abstract executions in which \co{r2} is greater than zero.
To see why not, consider that in any such execution
\co{P1()} would not store anything to \co{y};
it certainly would not store a positive value.
This means that the value loaded on line~8 would have to be \co{y}'s
initial value of zero, and so would the value stored to \co{x} on
line~9, and hence it would not be possible for the load on line~13
to get a positive value for \co{r2} after all.

@@

In other words, the compiler would have to preserve the dependency
ordering that the UB-based optimization would otherwise destroy.
It is tempting to consider using techniques
% Appendix~\ref{app:Aside on Undefined Behavior}
that prevent back-propagation of UB, but these are ineffective in this
case because the compiler can detect and exploit UB when
considering line~14 in isolation.

>>>>

Next, definitions:
\begin{description}
\item[coe]
	Coherence-order (external), which connects a store to any other store
	(in a different thread)
	to the same object that comes later in the object's
	modification order, i.e., that overwrites either the first store's
	value or some later value.
\item[fre]
	From-read (external), which connects a load to any store
	(in a different thread) to
	the same object that overwrites either the value that the load
	returned or some later value.
\item[rfe]
	Reads-from (external), which connects a store to any load
	(in a different thread) from
	the same object that returns the value stored.
	(Note: This means that the load retrieves the information
	written by the store, as opposed to retrieving the information
	from a different store that happened to write the same value.)
\end{description}

@@

Figure~\ref{fig:Propagation Delay and Temporal Properties of coe, fre, and rfe}
shows how propagation delay explains the temporal properties of coe,
fre, and rfe.
The upper portion of the figure shows an atemporal coe, in which
CPU~0 stores 1 to \co{x} after CPU~3 stores 2 but
the modification order of \co{x} is decided after the fact.
In this case, the value of 2 from CPU~3's store overwrites the value
of 1 from CPU~0's store, despite the fact that CPU~0's store
happened later in global time.
One way this could happen is if the cache line containing
\co{x} arrived at CPU~0 before arriving at CPU~3.
A more fanciful explanation is that both values arrived at an
interconnect at the same time, and the interconnect made an
arbitrary choice between the two.
Either way, the end result is that an earlier store can overwrite
a later store.

@@

The \co{-funsigned-char} causes the \co{char} type to be unsigned, which
overrides per-architecture defaults, some of which treat \co{char}
as signed and others as \co{unsigned}.
This choice prevents a class of bugs, and also allows the kernel to make
reliable use of the uppermost bit of variables of type \co{char}.
It also affects the definition of ``semantic dependency'' by changing
the arithmetic proporties of this type.
In theory, the standard could have specified the signedness of \co{char}
but the variety of existing practice in the 1980s prevented this.
Plus, optimizers were much less capable back then, so the signedness of
\co{char} was much more of a performance concern than it is today.

@@

However, this would mean that programs compiled in this way would not
be guaranteed to run correctly on weakly ordered systems that lack
ECC (or that lack precise exceptions).
In addition, this change is more strict than necessary, forbidding
optimizations that tooling could in fact analyze.
Therefore, the next two sections look at standardizing less severe
restrictions.

@@

@@

But if \co{x} is a nonvolatile atomic, there are no worries about
data races.
In theory, if the \co{do_something_with()} call was not present
the compiler could still make the above transformation without
worry, especially given that the value loaded from \co{x} is discarded in
the case where \co{limit} is less than or equal to zero.
Except that the data-race worry is replaced by a cache-miss worry;
after all, optimizations are supposed to speed things up,
not slow them down.
A compiler might be wise to adhere to the
same restriction that avoids data races for loads from non-atomic
variables in order to avoid potentially expensive cache misses.

@@

On line~14 of this listing, the computation \co{1 / (r2 <= 0)}
will yield one if \co{r2} is not positive and will result in UB
otherwise (a divide-by-zero error).
It may not be obvious, but the program is in fact UB-free---there are
no abstract executions in which \co{r2} is greater than zero.
To see why not, consider that in any such execution
\co{P1()} would not store anything to \co{y};
it certainly would not store a positive value.
This means that the value loaded on line~8 would have to be \co{y}'s
initial value of zero, and so would the value stored to \co{x} on
line~9, and hence it would not be possible for the load on line~13
to get a positive value for \co{r2} in the first place.

@@

In other words, the compiler would have to preserve the dependency-induced
ordering that the UB-based optimization would otherwise destroy.
It is tempting to consider using techniques
% Appendix~\ref{app:Aside on Undefined Behavior}
that prevent back-propagation of UB, but these are ineffective in this
case because the compiler can detect and exploit UB when
considering line~14 in isolation.
